Nice people or potential cooperators when keeping promises: an experimental and Bayesian account for two explanations

### 2019-08-08 article_english.tex
Updating manuscript.

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Nice guys or potential cooperators when keeping promises: an experimental and Bayesian account for two explanations},
            pdfauthor={Said Jiménez},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{Nice guys or potential cooperators when keeping promises: an
experimental and Bayesian account for two explanations}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Said Jiménez}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{July 2019}

@@

Human beings are social animals and we get advantages from being so.
Historically, belonging to a social group has provided protection and
guaranteed access to resources such as food, currently there has been
evidence that belonging to a social circle has a positive effect on
longevity, physical and mental health (Holt-Lunstad, 2018; Tough,
Siegrist, \& Fekete, 2017). The interrelation between social closeness
and cooperation is key to the formation of large groups of individuals
without a genetic relationship, such groups form the basis of
communities, societies, and nations, as well as probably constitute one
of the most fundamental conditions for human survival (Fehr \&
Fischbacher, 2003; Fehr \& Schurtenberger, 2018).

An important requirement for cooperation between individuals is
communication, a meta-analysis of 45 studies reported a large positive
effect (\(d = 1.01\)) of communication in cooperation regardless of the
communication medium (Balliet, 2010). In parallel, it is known that
after communication sessions between completely unknown individuals
there is an increase in subjective social closeness (Aron, Melinat,
Aron, Vallone, \& Bator, 1997). Thus, the findings suggest a positive
relationship between communication, cooperation, and social closeness.

One of the forms of communication that have received the most attention
in psychological game theory is the promise, in an interaction between
sender and recipient, it is believed that promises influence beliefs of
the recipient, generating trust and cooperation (Charness \& Dufwenberg,
2006; Vanberg, 2008). However, there are situations where trust is
betrayed, promises are broken or people deceive, for example,
approximately 11\% of people surveyed by the General Social Survey (GSS)
in 2018 responded that they have had sex with someone different to their
partner while they are married (Smith, Davern, Freese, \& Hout, 2018).
Another example, according to the Washington Post, President Trump has
said 10,796 false or misleading statements in 869 days in office
(Kessler, Rizzo, \& Kelly, 2019).

In the laboratory, these transgressions to trust have been explicitly
studied in experiments where subjects have incentives to lie or not keep
their promises. However, the vast majority of these studies in
economics, psychology and neurosciences have been conducted in people
who do NOT know each other (to name a few, Gneezy, 2005; Baumgartner,
Fischbacher, Feierabend, Lutz, \& Fehr, 2009; Baumgartner, Gianotti, \&
Knoch, 2013; Charness \& Dufwenberg, 2006; Fischbacher \& Föllmi-Heusi,
2013; Gneezy, Rockenbach, \& Serra-Garcia, 2013; Mazar, Amir, \& Ariely,
2008). In social psychology some studies address deception in close
interpersonal relationships, however, they use self-report measurements
(DePaulo, Ansfield, Kirkendol, \& Boden, 2004; DePaulo \& Kashy, 1998)
or rather explore the development of deception detection skills in
same-sex friends (Anderson, DePaulo, \& Ansfield, 2002). Despite the
widespread presence of interactions between human beings belonging to
the same group, it is not known how social closeness between subjects
can affect keeping or breaking promises.

In this manuscript, we explore the effects of three partners with
different levels of social closeness on keeping or breaking promises, as
well as on cooperation in a trust game. The above, we analyze it in the
light of the two main motivations that have been pointed out in the
literature regarding keeping promises (Baumgartner et al., 2009):

@@

In our study, subjects perform a standard trust game in pairs with three
phases: \emph{first}, the trustee makes a promise to pay half of his
earnings regardless of who his investor is; \emph{second}, the investor
receives the promise and decides if he invests his initial budget;
\emph{third}, in case the investor has given the budget, the trustee
faces the decision to pay or not to pay half of his earnings. To
evaluate the effect of social closeness, our subjects participate in the
role of trustee in front of three partners with different levels of
social closeness (zero, low and high) in the role of the investor: a
\emph{computer}, a \emph{stranger}, and a \emph{friend}.

The manipulations mentioned allow us to evaluate several hypotheses, the
first one was proposed a priori and is derived from a larger research
project (you can check the pre-registration of the hypothesis here:
\url{https://osf.io/u97fd}), while the following are exploratory:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Social closeness will reduce the decision of breaking the promise.
  According to the \emph{instrumental} motivation, we expect that
  subjects keep their promises with friends for the purpose of
  facilitating future cooperation, which can be extended even beyond the
  trials in the experiment. We also expect, although to a lesser extent,
  that the subjects keep promises to strangers, with the purpose of
  facilitating cooperation at least during the trials during the
  experiment. Finally, we anticipate that the participants break the
  promises to the computer because it is a partner without social
  closeness and they could not ensure cooperation in future trials. It
  should be noted that if the participants keep the promises to the
  computer, evidence would be given in favor of intrinsic motivation.
\item
  There will be an effect of social closeness on cooperation, regardless
  of promises. We expect that there is more probability of paying the
  \emph{friend} than the \emph{stranger} and more probability of paying
  the stranger than the \emph{computer}. According to the
  \emph{instrumental} motivation, the cooperation will be greater for
  partners with whom the subject anticipates greater cooperation in the
  future (friend \textgreater{} stranger \textgreater{} computer).
\item
  Finally, we will obtain two subsamples according to the cooperation
  rate of the subjects as was done in another study (Baumgartner et al.,
  2009). However, what in the aforementioned study was classified as a
  group of \emph{honest} and \emph{dishonest}, we will show that in our
  sample it is not supported because if the groups differ in their
  payment rate it will be due to the difference in their level of
  commitment expressed by promises.
\end{enumerate}

@@

We included 45 subjects (15 men), recruited from the National Autonomous
University of Mexico, the age range from 19 to 33 years and their
minimum educational level were bachelor's degrees. Subjects went to the
study with a friend considered \emph{close} by themselves, who fulfilled
the following characteristics: he was matched by sex, did not have a
family bond and was not a person with a sentimental or sexual
relationship. 30 subjects (15 men) performed the task within a Magnetic
Resonance Imaging (MRI) scanner, however, their image data is not
analyzed in the present work.

@@

All subjects performed an adaptation of the trust game with promises,
using hypothetical monetary rewards (Baumgartner et al., 2009). The task
was programmed in PsychoPy2 version 1.84.2 (Peirce et al., 2019; Peirce,
2008) and consisted of 24 trials between two players: trustee and
investor. The trustee originally has 0 Mexican pesos and the investor
has 2 pesos, the investor is presented with the opportunity to give his
money to the trustee or keep it. If he invests his money it is
multiplied by 5 pesos, so that the trustee has 10 pesos. Finally, the
trustee decides to pay half to the investor or keep the 10 pesos.

The structure described is repeated in 24 trials, however, in 4 of the
trials the trustee can send a promise to the investor, the promises were
that \emph{always}, \emph{mostly}, \emph{sometimes}, or \emph{never}
would pay back. Each promise was valid for three trials so that 12 of
the trials have the effect of the promise and the other 12 do not. It is
worth mentioning that the promises are not made to a particular partner,
but the subject expresses their commitment level to pay back regardless
of who their partners are in the next three trials in which the promise
is valid.

The variation with respect to the original task is that in our
experiment three partners were presented with three levels of social
closeness in the role of the investor: \emph{computer} (no closeness),
\emph{strange} (low closeness) and \emph{friend} (high closeness), while
the subject acts as the trustee. Each partner performs the role of
investor in 8 of 24 total trials, however, both the promises and social
closeness conditions were presented to the trustee in pseudorandom
order.

Since our main interest was the trustee's behavior, investors' decisions
were programmed \emph{a priori} to give their amount in 6 trials and in
2 did not. The covert story for all our subjects was that they would be
playing in real-time with their friend, the stranger (was told he would
be another same-sex unknown person) and the computer. A diagram of the
chronology of the experimental task with the duration of each phase in
seconds is shown in Figure 1, each box from left to right represents a
screen that was shown to the subjects sequentially. Section A\(_1\)
corresponds to an example of trials without promises and section A\(_2\)
to trials with promises.

Fixation phase consisted only of a period in which the subject paid
attention without performing any particular behavior, then,
\textbf{promises phase} in part A\(_1\) indicated to the subject that
the following three trials they could decide without the effect of the
promise, while in A\(_2\) the subject was asked to decide between
\emph{always}, \emph{mostly}, \emph{sometimes} or \emph{never} pay back.
This was followed by another fixation period, subsequently, trails in
both A\(_1\) and A\(_2\) continue in the same way.

In the \textbf{anticipation/assignation phase}, subjects were told who
their partner was for that trial (computer, stranger or friend) and was
given the message that their partner was making their decision. Later in
the \textbf{decision phase}, the subject was informed if his partner
invested his \$2 or not, he was reminded of his promise level (if they
were trails with promises) and, in case of his partner had invested, he
was asked to decide whether to pay back or not. Finally, in
\textbf{feedback phase} payments for that trial were shown and the
sequence was repeated.

@@

\caption{Trust game with promises}\label{fig:figA}
\end{figure}

@@

Subjects came with a same-sex friend considered close by him or her, it
was emphasized that they must not have a romantic or family relationship
with their friends, to try to exclude the effect on cooperation due to a
consanguineous relationship or sexual attraction. Also, to ensure that
the subjects and their companions had a similar degree of social
closeness to each other, in the laboratory both responded to the
``Inclusion of the Other in Self'' IOS scale (Aron et al., 1997) without
observing their partner responses. The scale consists of seven pairs of
circles that vary in the degree of overlap between them, the respondent
must select the pair of circles that best represents the subjective
closeness to his partner.

Subjects were trained in the task by two of the researchers, the game
was first explained verbally, and then the computerized interface was
presented to them on a portable computer. They were told that they would
make their decisions with four buttons on the computer keyboard, which
corresponded to the levels of promises and with only two of these to
make their decision to pay back or not. Friends heard the explanation,
they were told that they would make their decisions in a separate room
on another computer. Subjects performed between 3 and 6 practice tests
together with the researcher who exemplified the course of the game when
the investor gave his budget. In the practice trials, all the phases
indicated in Figure 1 were shown, however, the identity of the investors
was replaced by question marks.

When the subjects did not report doubts we proceeded to take them
without their companions to another room where they would make their
decisions on a laptop, or, with the Lumina Cedrus response system in the
case that it was a subject to which we also performed MRI. The subjects
began the task believing that they would really play with the three
partners of different social closeness. In the other room, another of
the researchers performed the \emph{debriefing} to the companion, they
were told that the purpose of the research was to know if their mere
presence had an effect on the subject's decisions, but that to simplify
the analysis their responses were programmed \emph{a priori} so your
participation ended there.

When the subject finished the task, \emph{debriefing} was also carried
out, he was told that the decisions of his partners had been programmed.
Although no participant exercised their right, both subjects and their
friends were told that they could withdraw their participation and
informed consent if they considered disagreeing with any of the
manipulations made by researchers.

@@

We analyzed 810 decisions about paying back or not (18 per subject), as
well as 180 promises (4 per subject), 69\% of the decisions, were pay
back, while the proportion of the promises selected were 5.4\%, 16.2\%,
43.8\% and 40.2\%, respectively for promises \emph{never},
\emph{sometimes}, \emph{mostly} and \emph{always}. Figure 2 shows the
payment proportion depending on the levels of promises and partners.

Breaking promises, that is, subject had promised \emph{always} and
subsequently decides not to pay back occurred only 12.4\% of trials (19
times they did not pay back of 153 when promised \emph{always}), of
which, 8.50\% (13/153) were trials with the computer, 3.92\% (6/153)
with the stranger and 0\% (0/153) with the friend. It is essential to
note that breaking the promise was not the most frequent decision, in
91.5\% of trials when they promised \emph{always} and had the computer
as a partner they decided to pay back.

@@

@@

All models presented below were programmed in R via the \textbf{brms}
package (Bürkner, 2017; R Core Team, 2019), which performs the inference
using sampling by Markov chain Monte Carlo through \textbf{Stan} (Stan
Development Team, 2018). For each model, the posteriors distributions of
all parameters were approximated with four chains of 2000 iterations
each, the first 1000 iterations of each chain were discarded (burning
period), for a total of 4000 post-burning samples. Models convergence
was evaluated through visual inspection of the chains and calculation of
the \(\hat{R}\) statistic, which for all parameters was 1, which is
interpreted as convergence. Vaguely informative prior distributions were
used for the parameters of interest in the models, which allows the data
to dominate the inference, also assumes as little as possible regarding
the nature of the phenomenon, which could be adequate for the current
state of evidence in the problem we are studying (McElreath, 2018).

To assess social closeness, 30 (15 men) of our subjects and their
friends responded to the IOS scale (Aron et al., 1997; Fareri \&
Delgado, 2014; Sip, Smith, Porcelli, Kar, \& Delgado, 2015), which
consist of seven pairs of circles that vary in the degree of overlap and
represent the social closeness that an individual perceives with respect
to the other. A Hierarchical and Cumulative Bayesian Model (See section:
Commitment expressed in the promises), supports the hypothesis that
there are no differences in subjective social closeness between subjects
and their friends, with a posterior evidence ratio of 2.74 in its favor.

@@

To evaluate the hypothesis that social closeness would reduce the
behavior of breaking the promise, we filtered all trials in which
subjects promised that they would \emph{always} pay back and then
decided not to pay. Subsequently, Bayesian inference was used to assess
the effect of partners on the decision to break the promise at the
individual and population level. For this purpose, a Hierarchical
Bayesian Model was carried out that assumes that the uncertainty in
partners effect on the decision to break the promise varies depending on
each individual, however, it also assumes that these variations belong
to common population distributions (Gelman \& Hill, 2006).

@@

The evidence ratio, which is the ratio between the posterior probability
of the mentioned hypotheses and their corresponding alternative
hypotheses was 7.47 and 8.64 respectively in favor of the previous
hypotheses. As can be seen in Figure 5 in the estimate for
\(\beta^{computer}\), when there is no social closeness we have a 95\%
posterior probability that the parameter for promise-breaking is between
4-38\%, although it is a small proportion, it clearly excludes the
probability that promises are not broken.

@@

Figure 4 shows the posterior distributions of the \(\beta\) coefficients
in the logit scale, the center line represents the median of the
distribution and the shaded area corresponds to the interval of 50\%
posterior probability. As can be seen, more than 95\% of the posterior
density of the population coefficients is greater than 0, which
indicates strong evidence of the effect of experimental conditions on
the decision to pay back, so that the presence of promises and the
stranger increase clearly the probability of paying, but the presence of
the friend is the condition that has the greatest effect on this
behavior. Posterior distributions clearly show that the probability of
cooperation increases as social closeness does.

Also, Figure 5 presents the posterior predictive distribution in
contrast to the proportion of decisions to each partner during the
promises phase, each panel corresponds to one of the first twelve
subjects. The responses on the dotted line would indicate that the
subject paid back random to that partner. Posterior predictive
distribution simulates observations of the model and compares them with
the actual data, it helps us to identify if the model is sufficiently
close to the process that generated the data (Lee \& Wagenmakers, 2014;
Schad, Betancourt, \& Vasishth, 2019). It can be seen that there is a
great correspondence between the subject's responses and predictions of
the model, even in cases where the model is ``wrong'' (for example,
subject 1), the observed response is in the range of a predicted
standard deviation, which gives credibility to estimates.

@@

@@

\leavevmode\hypertarget{ref-Kessler2019}{}%
Kessler, G., Rizzo, S., \& Kelly, M. (2019). President Trump has made
more than 10,000 false or misleading claims. Retrieved from
\href{https://www.washingtonpost.com/politics/2019/06/10/president-trump-has-made-false-or-misleading-claims-over-days/?utm\%7B/_\%7Dterm=.8565a16aa5e8}{https://www.washingtonpost.com/politics/2019/06/10/president-trump-has-made-false-or-misleading-claims-over-days/?utm\{\textbackslash{}\_\}term=.8565a16aa5e8}

@@

>>>>

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Nice people or potential cooperators when keeping promises: an experimental and Bayesian account for two explanations},
            pdfauthor={Said Jiménez},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{Nice people or potential cooperators when keeping promises: an
experimental and Bayesian account for two explanations}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Said Jiménez}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{August 2019}

@@

Human are social animals and we get survival advantages from being such.
Historically, belonging to a social group has provided protection and
guaranteed access to resources such as food, and currently, there has
been evidence that belonging to a social group has a positive effect on
longevity, physical and mental health (Holt-Lunstad, 2018; Tough,
Siegrist, \& Fekete, 2017). The relationship between social closeness
and cooperation is key to the formation of large groups of individuals
without a genetic relationship, and such groups form the basis of
communities, societies, and nations, arguably constituting one of the
most fundamental conditions for human survival (Fehr \& Fischbacher,
2003; Fehr \& Schurtenberger, 2018). An important requirement for
cooperation between individuals is communication. A meta-analysis of 45
studies reported a large positive effect (\(d = 1.01\)) of communication
in cooperation regardless of the communication medium (Balliet, 2010).
In parallel, it is known that after communication sessions between
completely unknown individuals there is an increase in subjective social
closeness (Aron, Melinat, Aron, Vallone, \& Bator, 1997). Thus, these
findings suggest a positive relationship between communication,
cooperation, and social closeness.

One of the forms of communication that have received the most attention
in psychological game theory is ``the promise''. In an interaction
between sender and recipient, it is believed that promises influence
beliefs of the recipient, generating trust and cooperation (Charness \&
Dufwenberg, 2006; Vanberg, 2008). However, there are situations where
trust is betrayed, promises are broken or people deceived. The General
Social Survey (GSS) in 2018 showed that approximately 11\% of people
surveyed responded that they have had sex with someone different to
their partner while they are married (Smith, Davern, Freese, \& Hout,
2018). In economic and social terms, the United Nations Organization has
stressed that deception (corruption) has a global cost of at least 5
percent of the annual gross domestic product (GDP), and it is a
generator of conflict and instability in all nations (UN, 2018).

Despite the widespread presence of interactions between human belonging
to the same group, it is not known how social closeness between
participants can affect keeping or breaking promises. This is important
because it has been pointed out that lies and betrayals of trust occur
daily in close relationships, such as friends, workmates, or classmates.
And, their consequences can be serious such as loss of the relationship,
work or damage to reputation (DePaulo, Ansfield, Kirkendol, \& Boden,
2004). In the laboratory, these transgressions to trust have been
explicitly studied in experiments where subjects have incentives to lie
or not keep their promises. However, the vast majority of these studies
have been conducted in people who do NOT know each other (Baumgartner,
Fischbacher, Feierabend, Lutz, \& Fehr, 2009; Baumgartner, Gianotti, \&
Knoch, 2013; Charness \& Dufwenberg, 2006; Fischbacher \& Föllmi-Heusi,
2013; Gneezy, 2005; Gneezy, Rockenbach, \& Serra-Garcia, 2013; Mazar,
Amir, \& Ariely, 2008). Some studies address deception in close
interpersonal relationships using self-report measurements, however,
they have limitations due to social desirability bias (DePaulo et al.,
2004; DePaulo \& Kashy, 1998) or rather explore the development of
deception detection skills in same-sex friends (Anderson, DePaulo, \&
Ansfield, 2002).

In this study, we wanted to understand the effects of three partners
with different levels of social closeness on keeping or breaking
promises, as well as on cooperation, using an experimental trust game.
We analyzed it in light of the two main motivations that have been
pointed out in the literature regarding keeping promises which are:
instrumental and intrinsic (Baumgartner et al., 2009):

@@

In our study, subjects perform a standard trust game in pairs with three
phases: \emph{first}, the trustee makes a promise to pay half of his
earnings regardless of who his investor is; \emph{second}, the investor
receives the promise and decides if he invests his initial budget;
\emph{third}, in case the investor has given the budget, the trustee
faces the decision to pay or not to pay half of his earnings. To
evaluate the effect of social closeness, our subjects participate in the
role of trustee in front of three partners with different levels of
social closeness (null, low and high) in the role of the investor: a
computer, a stranger, and a friend. The manipulations mentioned allow us
to evaluate several hypotheses, the first one was proposed a priori and
is derived from a larger research project (you can check the
preregistration of the hypothesis here: \url{https://osf.io/u97fd}),
while the following are exploratory:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Social closeness will reduce the decision of breaking the promise.
  According to the \emph{instrumental} motivation, we expect that
  subjects keep their promises with friends for the purpose of
  facilitating future cooperation, which can be extended even beyond the
  trials in the experiment. We also expect, although to a lesser extent,
  that the subjects keep promises to strangers, with the purpose of
  facilitating cooperation at least during the trials during the
  experiment. Finally, we anticipate that the participants break the
  promises to the computer because it is a partner without social
  closeness and they could not ensure cooperation in future trials. It
  should be noted that if the participants keep the promises to the
  computer, evidence would be given in favor of \emph{intrinsic}
  motivation.
\item
  There will be an effect of social closeness on cooperation, regardless
  of promises. We expect that there is more probability of paying the
  friend than the stranger and more probability of paying the stranger
  than the computer. According to the \emph{instrumental motivation},
  the cooperation will be greater for partners with whom the subject
  anticipates greater cooperation in the future (friend \textgreater{}
  stranger \textgreater{} computer).
\item
  Finally, we will obtain two subsamples according to the cooperation
  rate of the subjects as was done in another study (Baumgartner et al.,
  2009). However, what in the aforementioned study was classified as a
  group of ``honest'' and ``dishonest'', we will show that in our sample
  it is not supported because if the groups differ in their payment rate
  it will be due to the difference in their level of commitment
  expressed by promises.
\end{enumerate}

@@

We recruited 45 subjects (15 male) from the National Autonomous
University of Mexico, the age range from 19 to 33 years and their
minimum educational level were bachelor. Subjects were asked to come to
the study with a ``close'' friend, who fulfilled the following
characteristics: 1) he/she was matched by sex, 2) did not have a family
bond and 3) was not a person with a sentimental or sexual relationship.
From this sample, 30 subjects (15 men) performed the task in a magnetic
resonance imaging (MRI) scanner, however, their image data was not
analyzed for the present work.

@@

All subjects performed an adaptation of the trust game with promises,
using hypothetical monetary rewards (Baumgartner et al., 2009). The
variation with respect to the original task is that in our experiment
three partners were presented with three levels of social closeness in
the role of the investor: computer (no closeness), strange (low
closeness) and friend (high closeness), while the participant acted as
the trustee. The task was programmed in PsychoPy2 version 1.84.2 (Peirce
et al., 2019; Peirce, 2008) and consisted of 24 trials between two
players: trustee and investor. The trustee originally has 0 Mexican
pesos and the investor has 2 pesos, the investor is presented with the
opportunity to give his money to the trustee or keep it. If investor
gives his money, it is multiplied by 5 pesos, so that the trustee has 10
pesos. Finally, the trustee decides to pay half to the investor or keep
the 10 pesos. The structure described is repeated in 24 trials, however,
in 4 of the trials the trustee can send a promise to the investor, the
promises were that always, mostly, sometimes, or never would pay back.
Each promise was valid for three trials so that 12 of the trials have
the effect of the promise and the other 12 do not. Each partner performs
the role of investor in 8 of 24 total trials, however, both the promises
and social closeness conditions were presented to the trustee in
pseudorandom order.

Since our main interest was the trustee's behavior, investors' decisions
were programmed \emph{a priori} to give their amount in 6 trials and in
2 did not randomly. The covert story for all our subjects was that they
would be playing in real-time with their friend (whom they brought to
the study), the stranger (was told he would be another same-sex unknown
person) and the computer. After the study ended, the participants were
debriefed about our own deception. A diagram of the chronology of the
experimental task with the duration of each phase in seconds is shown in
Figure 1.

@@

\caption{Trust game with promises, each box from left to right represents a screen that was shown to the subjects sequentially. Section A1 corresponds to an example of trials without promises and section A2 to trials with promises.}\label{fig:figA}
\end{figure}

Fixation phase consisted only of a period in which the subject paid
attention without performing any particular behavior. Then,
\textbf{promises phase} in part A1 indicated that in the following three
trials they could decide without the effect of the promise. In A2, the
subject was asked to decide between 1) always, 2) mostly, 3) sometimes
or 4) never pay back. This pattern was repeated. In the
\textbf{anticipation/assignation phase}, subjects were told who their
partner was for that trial (computer, stranger or friend) and was given
the message that their partner was making their decision. Later in the
\textbf{decision phase}, the subject was informed if his partner
invested his \$2 or not, he was reminded of his promise level (if they
were trails with promises) and, in case the partner had invested, he was
asked to decide whether to pay back or not. Finally, payments for each
trial were shown in the \textbf{feedback phase} and the sequence was
repeated until the task finished.

@@

Subjects came to the lab with a same-sex friend considered close by him
or her. It was emphasized that they must not have a romantic or family
relationship with their friends, to try to exclude the effect on
cooperation due to a consanguineous relationship or sexual attraction.
Also, to ensure that the subjects and their companions had a similar
degree of social closeness to each other, both responded the ``Inclusion
of the Other in Self'' IOS scale (Aron et al., 1997) without observing
their partner responses. The scale consists of seven pairs of circles
that vary in the degree of overlap between them, the respondent must
select the pair of circles that best represents the subjective closeness
to his partner. Subjects were trained in the main task by two
researchers, the game was first explained verbally, and then the
computer interface was presented on a portable computer. They were told
that they would make their decisions with four buttons on the computer
keyboard, which corresponded to the levels of promises and with only two
of these to make their decision to pay back or not. Friends heard the
explanation, they were told that they would make their decisions in a
separate room on another computer even though they were not. Subjects
performed between 3 and 6 practice tests together with the researcher
who exemplified the course of the game when the investor gave his
budget. When the subjects were ready, we proceeded to take them without
their companions to another room (or MRI scanner it that was the case)
where they would perform the task.The subjects began the task believing
that they would really play with the three partners of different social
closeness. In the other room, we debriefed the companion about the
deception. When the subject finished the task, they were debriefed.
Although no participant exercised their right, both subjects and their
friends were told that they could withdraw their participation and
informed consent if they considered disagreeing with any of the
manipulations made by the researchers.

\hypertarget{bayesian-modeling}{%
\subsection{Bayesian Modeling}\label{bayesian-modeling}}

All models presented below were programmed in R via the \textbf{brms}
package (Bürkner, 2017; R Core Team, 2019), which performs the inference
using sampling by Markov chain Monte Carlo through \textbf{Stan} (Stan
Development Team, 2018). For each model, the posteriors distributions of
all parameters were approximated with four chains of 2000 iterations
each, the first 1000 iterations of each chain were discarded (burning
period), for a total of 4000 post burning samples. Models convergence
was evaluated through visual inspection of the chains and calculation of
the \(\hat{R}\) statistic, which for all parameters was 1, that can be
interpreted as convergence. Vaguely informative prior distributions were
used for the parameters of interest in the models, which allows the data
to dominate the inference, also assumes as little as possible regarding
the nature of the phenomenon, which could be adequate for the current
state of evidence in the problem we are studying (McElreath, 2018).

@@

We analyzed 810 decisions about paying back or not (18 per subject), as
well as 180 promises (4 per subject). 69\% of the decisions, were pay
back, while the proportion of the promises selected were never = 5.4\%,
sometimes = 16.2\%, mostly = 43.8\% and always = 40.2\%. Figure 2 shows
the payment proportion depending on the levels of promises and partners.
Breaking promises, meaning the subject had promised always and
subsequently decides not to pay back, occurred only in 12.4\% of trials
(19 times they did not pay back of 153 when promised always), of which,
8.50\% (13/153) were trials with the computer, 3.92\% (6/153) with the
stranger and 0\% (0/153) with the friend. It is essential to note that
breaking the promise was not the most frequent decision, when they
promised always and had the computer as a partner, they decided to pay
back in 91.5\% of trials.

@@

To assess social closeness, 30 (15 men) of our subjects and their
friends responded to the IOS scale (Aron et al., 1997; Fareri \&
Delgado, 2014; Sip, Smith, Porcelli, Kar, \& Delgado, 2015), which
consists of seven pairs of circles that vary in the degree of overlap
and represent the social closeness that an individual perceives with
respect to the other. A Hierarchical and Cumulative Bayesian Model (See
section: Commitment expressed in the promises), supports the hypothesis
that there are no differences in subjective social closeness between
subjects and their friends, with a posterior evidence ratio of 2.74 in
its favor.

@@

@@

To evaluate the hypothesis that social closeness would reduce the
behavior of breaking the promise, we filtered all trials in which
subjects promised that they would always pay back and then decided not
to pay. Subsequently, Bayesian inference was used to assess the effect
of partners on the decision to break the promise at the individual and
population level. For this purpose, a Hierarchical Bayesian Model was
carried out that assumes that the uncertainty in partners effect on the
decision to break the promise varies depending on each individual,
however, it also assumes that these variations belong to common
population distributions (Gelman \& Hill, 2006).

@@

The evidence ratio, which is the ratio between the posterior probability
of the mentioned hypotheses and their corresponding alternative
hypotheses was 7.47 and 8.64 respectively in favor of the previous
hypotheses. As seen in Figure 5 in the estimate for
\(\beta^{computer}\), no social closeness, we have a 95\% posterior
probability that the parameter for promise-breaking is between 4-38\%,
although it is a small proportion, it clearly excludes the probability
that promises are not broken.

@@

Figure 4 shows the posterior distributions of the \(\beta\) coefficients
in the logit scale, the center line represents the median of the
distribution and the shaded area corresponds to the interval of 50\%
posterior probability. As shown, more than 95\% of the posterior density
of the population coefficients is greater than 0, which shows strong
evidence of the effect of experimental conditions on the decision to pay
back, although the presence of promises and the stranger clearly
increase the probability of paying, the presence of the friend is the
condition that has the greatest effect on this behavior. Posterior
distributions clearly show that the probability of cooperation increases
as social closeness does.

@@

Figure 5 shows the posterior predictive distribution in contrast to the
proportion of decisions to each partner during the promises phase, and
each panel corresponds to one of the first twelve subjects. The
responses on the dotted line would indicate that the subject paid back
random to that partner. Posterior predictive distribution simulates
observations of the model and compares them with the actual data, it
helps us to identify if the model is sufficiently close to the process
that generated the data (Lee \& Wagenmakers, 2014; Schad, Betancourt, \&
Vasishth, 2019). It can be seen that there is a great correspondence
between the subject's responses and predictions of the model. Even in
cases where the model is ``wrong'' (for example, subject 1), the
observed response is in the range of a predicted standard deviation,
which gives credibility to estimates.

@@

@@

\leavevmode\hypertarget{ref-UN2018}{}%
UN. (2018). Global Cost of Corruption at Least 5 Per Cent of World Gross
Domestic Product, Secretary-General Tells Security Council, Citing World
Economic Forum Data. Retrieved from
\url{https://www.un.org/press/en/2018/sc13493.doc.htm}
